**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-10 Manuscripts/Manuscript_AM_v1.Rmd
updated

bibliography: refs.bib
output:
  word_document:
    reference_docx: docx_template.docx
  pdf_document: 
    fig_caption: yes
  css: style.css
  html_document: default
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

$$ y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}$$
Where

@@

#### 1.2 Constant correlation

Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
#### 1.1 Constant correlation

While the _repeated measures analysis of variance_ (rm-ANOVA) is the norm in the biomedical community to analyze longitudinal data, two frequent situations that arise in longitudinal studies limit its applicability. 
Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)

One of the key assumptions for rm-ANOVA is the constant correlation among measurements. This frequently not the case as the correlation reduces as the time interval between two measurements increases [@ugrinowitsch2004]. From a practical

#### 1.2 the case of LMEMs using splines and how they work and how they are better than rm-ANOVA

>>>>

bibliography: refs.bib
output:
  pdf_document: 
    fig_caption: yes
  html_document: default
  css: style.css
  word_document:
    reference_docx: docx_template.docx
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

\begin{equation}
(\#eq:labelANOVA)
y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}
\end{equation}

Where

@@

_Question: How can one make a plot that tests the "limits" of the model, i.e how much "wiggliness" can this model accomodate? Make a plot to show the behavior of the model_

#### 1.2 Covariance and correlation in rm-ANOVA and LMEMs

In a longitudinal study, the fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model. Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject. The latter is then taken in to account by a _covariance_ structure [@wolfinger1996].  In this case, _covariance_ can be defined as the dependency  between two different values, e.g. if higher values of a variable correspond with higher values of another value, the covariance is positive.  Although not immediately apparent, the reason for this specification arises from the fact that the model from \@ref(eq:labelANOVA) is re-written in matrix form for computational purposes (For an in-depth analysis see [@west2014;@weiss2005]).

For an rm-ANOVA analysis, the _variance-covariance matrix_ or the matrix that specifies the relationships between the different repeated measures, needs to have a specific construction known as _compound symmetry_ (which is commonly known also as "sphericity" or "circularity"). This assumes that variance of observations and the correlation between observations are constant,and when this is not the case, adjustments need to be made  [@weiss2005;@geisser1958;@huynh1976].However, it has been shown when the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].This assumption is frequently unjustified as the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].

In the case of LMEMs, they have the advantage of allowing different structures of the variance-covariance matrix [@pinheiro2006].

*make plot that explains between and within correlation*

#### 1.3 Missing observations

In a longitudinal study, missing observations are an important issue that arises frequently. In biomedical research, this situation is not unusual and can be caused by different reasons in the context of each study, but they are typically beyond the control of the investigator [molenberghs2004].statistical treatment of missing observations is particularly important for longitudinal studies. 
*talk about missing at random, mcar etc*

One of the key assumptions for rm-ANOVA is the constant correlation among measurements. This frequently not the case as the correlation reduces as the time interval between two measurements increases []. From a practical

#### 1.2 the case of GMEMs using splines and how they work and how they are better than rm-ANOVA
